Data migration under a Changing Schema

### 2022-12-27 2022 Migration/articleMigration.tex
substitutions for relations added

\subsection{Data migration}
   Sections~\ref{sct:Information Systems} and~\ref{sct:Example existing IS} both show a complete information system.
   This section studies the transition from an existing system to a desired one, involving a data migration.

   To migrate system $\infsys$ to $\infsys'$, we take the following steps:
\begin{enumerate}
   \item Let the existing system $\infsys$ have state $\dataset$ and schema $\schema$.
         A developer defines a desired system $\infsys'$, which contains a dataset of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.
         After the migration, the system will contain the triples from $\dataset$, to preserve the existing data.
         The developer may want to add some new data into the system,
         for instance to initialize new features on the moment of transition (MoT).
         She specifies such data in $\dataset'$.
         If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
   \item The generator generates a migration script $\migrsys$ based on $\schema$ and $\infsys'$
         to migrate the data and define the semantics during the migration phase as far as a generator can reasonably predict.
         The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the desired dataset.
   \item A migration engineer changes $\id{migrationrules}$ to accommodate the non-standard requirements of the migration,
         yielding migration script $\migrsys'$.
         This schema also contains data structures and rules that are needed for the duration of the migration phase.
         The migration engineer does not alter the existing system $\infsys$ nor the specification of the desired system $\infsys'$ that the developer has produced.
         As a result, the migration engineer can always restore the existing system because it is part of the migration system.
   \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
         while the data of the existing system remains intact.
         This can be realized by configuring the ingress system to open the route to the desired system,
         while keeping the existing system accessible.
         This event marks the MoT.
         From this point onward, both the existing functionality and the desired functionality are available to users as illustrated in figure~\ref{fig:migration phase}.
   \item To ensure that all data is migrated before the the migration engineer makes the migration irreversible,
         we require that the rules of the the migration system $\rules_{\migrsys'}$ are at rest before the migration engineer takes down the existing system.
         Users (i.e.\ the business) must do the work to satisfy the manually enforced rules.
         Ampersand guarantees that enforce rules will eventually be satisfied.
   \item To give the business an opportunity to evaluate the change,
         we can require a go/no-go decision before the the migration engineer makes the migration irreversible.
   \item In case of a go-decision, the migration engineer can disable the undo-facility as soon as the migration system is at rest.
         In case of a no-go, the migration engineer restores the existing system.
         This event marks the moment of no return (MoN).
   \item The system retains the data, the schema, and the functionality that is covered by $\schema'$, dropping everything else.
         This establishes the post-migration situation (figure~\ref{fig:post-migration}).
\end{enumerate}

   Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
\begin{definition}[disjoint union of datasets]
\begin{eqnarray}
   \rlap{$\pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}$}\notag\\
   &=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\notag\\
      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
\end{eqnarray}
\end{definition}

@@

To implement the disjoint union, we need to relabel 
\begin{definition}[relabel concepts]
   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
\end{definition}
\begin{definition}[relabel concepts in triples]
   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
\end{definition}
\begin{definition}[relabel concepts in relations]
   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
\end{definition}
\begin{definition}[relabel concepts in $\inst$]
   \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\ \inst\ A\]
\end{definition}
\begin{definition}[relabel concepts in datasets]
   \[\begin{array}{rcl}
      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
   \end{array}\]
\end{definition}
\begin{definition}[relabel concepts in rules]
   \[\sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}\]
\end{definition}
\begin{definition}[relabel concepts in schemas]
   \begin{eqnarray}
      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
      &=&\begin{array}[t]{l@{}l}
         \la&\{\subst{C}{D}{c}\mid c\in\concepts\}\\
         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
         \end{array}
   \end{eqnarray}
\end{definition}

>>>>

\subsection{Process of Data Migration}
   Sections~\ref{sct:Information Systems} and~\ref{sct:Example existing IS} introduce a single information system.
   The transition from an existing system to the desired one is the topic of this section.
   Let us first sketch the steps to migrate system $\infsys$ to $\infsys'$ informally.
\begin{enumerate}
   \item Let the existing system $\infsys$ have state $\dataset$ and schema $\schema$.
         A developer defines a desired system $\infsys'$, which contains a dataset of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.
         After the migration, the system will contain the triples from $\dataset$, to preserve the existing data.
         The developer may want to add some new data into the system,
         for instance to initialize new features on the moment of transition (MoT).
         She specifies such data in $\dataset'$.
         If $\schema$ equals $\schema'$, we call the migration trivial because a data migration is not necessary.
   \item The generator generates a migration script $\migrsys$ based on $\schema$ and $\infsys'$
         to migrate the data and define the semantics during the migration phase as far as a generator can reasonably predict.
         It generates a schema $\id{migrationrules}$, which contains the enforce rules that migrate existing data to the desired dataset.
   \item A migration engineer changes $\id{migrationrules}$ to accommodate the non-standard requirements of the migration,
         yielding migration script $\migrsys'$.
         This script also contains data structures and rules that are needed for the duration of the migration phase.
         The migration engineer does not alter the existing system $\infsys$ nor the specification of the desired system $\infsys'$ that the developer has produced.
         As a result, the migration engineer can always restore the existing system.
   \item The production environment switches instantly from $\infsys$ to $\migrsys'$.
         This can be realized by configuring the ingress system to open the route to the migration system and the desired system,
         while keeping the existing system accessible.
         This event marks the MoT.
         From this point onward, both the existing functionality and the desired functionality are available to users as illustrated in figure~\ref{fig:migration phase}.
   \item To ensure that all data is migrated before the the migration engineer makes the migration irreversible,
         we require that every rule from $\rules_{\migrsys'}$ is satisfied before the migration engineer takes down the existing system.
         Users (i.e.\ the business) must do the work to satisfy the manually enforced rules.
         Ampersand guarantees that enforce rules will be satisfied.
   \item To give the business an opportunity to evaluate the change,
         we can require a go/no-go decision before the the migration engineer makes the migration irreversible.
         In case of a go-decision, the migration engineer can disable the undo-facility as soon as the migration rules are satisfied.
         In case of a no-go, the migration engineer restores the existing system.
         This event marks the moment of no return (MoN).
   \item The system retains the data, the schema, and the functionality that is covered by $\schema'$ by dropping everything else.
         This establishes the post-migration situation (figure~\ref{fig:post-migration}).
\end{enumerate}

\subsection{Formalization}
   Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
\begin{definition}[disjoint union of datasets]
\begin{eqnarray}
   \rlap{$\pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}$}\notag\\
   &=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\notag\\
      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
\end{eqnarray}
\end{definition}

@@

To implement the disjoint union, we need to relabel 
\begin{definition}[relabel concepts]
   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
\end{definition}
\begin{definition}[relabel concepts in triples]
   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
\end{definition}
\begin{definition}[relabel concepts in relations]
   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
\end{definition}
\begin{definition}[relabel concepts in $\inst$]
   \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\ \inst\ A\]
\end{definition}
\begin{definition}[relabel concepts in datasets]
   \[\begin{array}{rcl}
      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
   \end{array}\]
\end{definition}
\begin{definition}[relabel concepts in rules]
   \[\sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}\]
\end{definition}
\begin{definition}[relabel concepts in schemas]
   \begin{eqnarray}
      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
      &=&\begin{array}[t]{l@{}l}
         \la&\{\subst{C}{D}{c}\mid c\in\concepts\}\\
         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
         \end{array}
   \end{eqnarray}
\end{definition}
\begin{definition}[relabel relations]
   \[\subst{\declare{n}{A}{B}}{r}{{\declare{m}{C}{D}}}\ =\ \text{\bf if}\ n=m\wedge A=C\wedge B=D\ \text{\bf then}\ r\ \text{\bf else}\ \declare{m}{C}{D}\]
\end{definition}
\begin{definition}[relabel relations in triples]
   \[\subst{p}{q}{\triple{a}{r}{b}} = \triple{a}{\subst{p}{q}{r}}{b}\]
\end{definition}
\begin{definition}[relabel relations in datasets]
   \[\begin{array}{rcl}
      \rlap{$\subst{p}{q}{\pair{\triples}{\inst}}$}\\
      &=&\pair{\{\subst{p}{q}{t}\mid t\in\triples\}}{\inst}
   \end{array}\]
\end{definition}
\begin{definition}[relabel relations in rules]
   \[\sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{p}{q}{u}}{\subst{p}{q}{\dataset}}\]
\end{definition}
\begin{definition}[relabel relations in schemas]
   \begin{eqnarray}
      \rlap{$\subst{\declare{n}{A}{B}}{{\declare{m}{C}{D}}}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
      &=&\begin{array}[t]{l@{}l}
         \la&\{\subst{B}{D}{{\subst{A}{C}{c}}}\mid c\in\concepts\}\\
         ,&\{\subst{B}{D}{{\subst{A}{C}{{\subst{\declare{n}{A}{B}}{{\declare{m}{C}{D}}}{r}}}}}\mid r\in\rels\}\\
         ,&\{\subst{B}{D}{{\subst{A}{C}{{\subst{\declare{n}{A}{B}}{{\declare{m}{C}{D}}}{u}}}}}\mid u\in\rules\}\ \ra\notag
         \end{array}
   \end{eqnarray}
\end{definition}
